Keep your Laziness in Check

### 2018-02-13 paper/main.tex

Moreover, it is important to be able to distinguish whether such a
datatype follows to a desired laziness behavior. For instance, the
standard functional solution for dynamic programming problems
crucially uses laziness. In particular, consider the famous knapsack
problem.
%
The inputs of the problem are two sequences of numbers $w_i$ and $v_i$
where $i = 1, \dots, n$, and each of $w_i$ and $v_i$ specify the
weight and the value of the $i$th item. The solution presents the
maximum value achievable using a subset of the items such that their
combined weight is under a limit $W$.

The usual dynamic programming approach for knapsack maintains a
table $OPT[][]$, where the entry $OPT[j][k]$ keeps the best value
achievable using only items $1\dots j$ and a weight limit of $k$. The
table satisfies the following recurrence relation: $$OPT[j][k]
= \max(OPT[j-1][w], w_j + OPT[j-1][k-w_j])$$

To efficiently compute this table using imperative langauges, we look
for shared sub-problems of the original problem, and design a path
along the solution table that maximizes sharing while filling out the
answers.
%
However, lazy evaluation automatically finds this path if we
use \lk{Data.Map.Lazy} and setup the solution table with the
recurrence relation \hzh{We can probably make this code prettier}
\begin{inlinecode}
ws, vs :: [Int]
n,  w  :: Int

@@

\leo{Is this true? Where would the infinite loop be? there is a bottom to the recursion.
I thought the problem was that it computes everything when it doesn't need to}
Using the lazy variant of \lk{Map} here is critical for the correctness
of \lk{solution}. If we used \lk{Data.Map.Strict} instead, then \lk{fromList}
would have eagerly evaluated the \lk{if} expression. The \lk{if} expression
contains value-level recursion to itself, and this would manifest as an infinite
loop at runtime. This is an incredibly subtle bug related to laziness.

As these examples show, the power of laziness puts great
responsibility on programmers' shoulders. An incorrect implementation
can either behave functionally the same but with inferior performance,
or it can completely break down! Reasoning about lazy programs is also
notoriously hard because predicting when a lazy value gets evaluated
is difficult.

@@

>>>>

Moreover, it is important to be able to distinguish whether such a
datatype follows to a desired laziness behavior. For instance, the
standard functional solution for dynamic programming problems
crucially uses laziness. In particular, consider the famous knapsack
problem.
%
The inputs of the problem are two sequences of numbers $w_i$ and $v_i$
where $i = 1, \dots, n$, and each of $w_i$ and $v_i$ specify the
weight and the value of the $i$th item. The solution presents the
maximum value achievable using a subset of the items such that their
combined weight is under a limit $W$.
%
The usual dynamic programming approach for knapsack maintains a
table $OPT[][]$, where the entry $OPT[j][k]$ keeps the best value
achievable using only items $1\dots j$ and a weight limit of $k$. The
table satisfies the following recurrence relation: $$OPT[j][k]
= \max(OPT[j-1][w], w_j + OPT[j-1][k-w_j])$$

To efficiently compute this table using imperative langauges, we look
for shared sub-problems of the original problem, and design a path
along the solution table that maximizes sharing while filling out the
answers.
%
However, lazy evaluation automatically finds this path if we use lazy
maps to represent the solution table according to the recurrence
relation \hzh{We can probably make this code prettier}
\begin{inlinecode}
ws, vs :: [Int]
n,  w  :: Int

@@

To query the optimal achievable values from the solution table, we can
just index into the map. When an entry from the map is looked up at
the first time, a sequence of evaluations is triggered, and their
results memoized in the lazy map automatically.

\leo{Is this true? Where would the infinite loop be? there is a bottom to the recursion.
I thought the problem was that it computes everything when it doesn't need to}
Using the lazy variant of \lk{Map} here is critical for the correctness
of \lk{solution}.
However, using the lazy variant of \lk{Map} here is critical for the
correctness of \lk{solution}. If we used a map strict in its values
instead, then \lk{fromList} would have eagerly evaluated the \lk{if}
expression. The \lk{if} expression contains value-level recursion to
the solution map itself, and this would manifest as an infinite loop
at runtime. This is an incredibly subtle bug related to laziness.

As these examples show, the power of laziness puts great
responsibility on programmers' shoulders. An incorrect implementation
can either behave functionally the same but with inferior performance,
or it can completely break down! Reasoning about lazy programs is also
notoriously hard --- the programmer needs to have a mental model of the
order of evaluation of the code, and this order is not always obvious
from the syntax of the code.

@@

In this section, we will discuss how to specify the laziness behaviors
of Haskell functions through running examples of purely functional
maps from integers to integers. The map operations of interest
are \lk{insert} and \lk{lookup} with the following types:
\begin{inlinecode}
insert :: IntMap -> Int -> Int -> IntMap Int
lookup :: IntMap -> Int        -> Maybe  Int
\end{inlinecode}
%
The \lk{IntMap} can be implemented with a few different laziness
flavors. We can choose a binary search tree representation for
efficient lookups, or we can choose an association list representation
for easy inserts. The binary search tree representation necessarily
evaluates the keys of the map to order the keys, but it can be lazy in
the values. On the other hand, the association list representation can
be lazy in the keys as well. To summarise, we have maps that are
strict in both keys and values, maps that are strict in keys but lazy
in values, and maps that are fully lazy in both keys and values.

In Haskell, triggering the evaluation of a lazy function requires
exerting some demand on its result. Such demands can be pattern
matches, or further computations that depend on the result. In typical
Haskell programming, such demands are exerted by the contextual code
surrounding the result of a lazy function. We will use the
phrase \textit{demand context} to refer to functions that causes
evaluation of lazy values.

An evaluation of a lazy data structure can partially explore the lazy
structure's shape. For example, \lk{lookup} on an association list
based map evaluates the association list until it finds the first
matching key. If such a key is halfway inside the association list,
then the values past this key will not be evaluated in this operation.
%
To describe how much of a data structure is evaluated, we will
use \textit{demand types}. A \textit{demand type} is almost isomorphic
to the corresponding data type, except we will wrap all the
constructor arguments in a \lk{Thunk} to represent the possible lack
of evaluation. For example, consider the \lk{Tree} type and
its \textit{demand type} \lk{TreeD}.
